Predicting Electoral Outcomes: Bayesian Modeling of Voter Support Trends for Kamala Harris and Donald Trump in the 2024 Presidential Election

### 2024-11-04 paper/paper.qmd
add comment in script and more graphs in data section

```{r}
#| include: false
#| warning: false
#| message: false
library(dplyr)
library(tidyr)
library(knitr)
library(rstanarm)
library(ggplot2)
library(tidyverse)
library(arrow)
```

@@

The analysis utilizes FiveThirtyEight's dataset of national presidential general election polls [@fivethirtyeight]. Following the approach outlined in [@tellingstories], we aim to predict the election outcome based on this polling data. The analyses were conducted in R [@citeR], utilizing several packages to streamline the process. The tidyverse packages [@citetidyverse] were essential for data simulation and preliminary testing. Initially, the raw data was downloaded using the arrow package [@citearrow], allowing for efficient handling of large datasets. The data cleaning process employed the dplyr [@citedplyr] and tidyr [@citetidyr] packages to reshape and organize the data effectively.

Model construction involved the rstanarm [@citerstanarm] package for Bayesian regression modeling, complemented by the splines package [@citesplines] to account for non-linear relationships in the data. The results were summarized using the modelsummary [@citemodelsummary] package, while visualizations of the findings were created with the ggplot2 [@citeggplot] package, ensuring clear communication of the model's outputs.

@@

## Predictors Explanation
- **state**: The U.S. state where the poll was conducted, which allows for state-by-state analysis and comparison of support levels.

- **end date**: The date the poll concluded, marking the end of data collection for that specific poll.

- **sample size**: The number of respondents in the poll, indicating the scope and potential statistical reliability of the results.

@@

@@

- **pollscore**: A specific rating of the individual poll, reflecting additional factors that impact the poll's quality and reliability.

- **methodology**: The method used by the pollster to conduct the poll, which may affect the reliability and interpretation of the results.

- **transparency score**: A measure of how openly the pollster reports their methodology and results. Higher scores suggest greater transparency and reliability.

@@

# Visualization
ggplot(president_polls_cleaned_data, aes(x = end_date, y = pct, color = candidate_name)) +
  geom_point(aes(shape = state), alpha = 0.5) +  # Use state as shape
  geom_line(
    data = pred_summary,
    aes(x = end_date, y = pred_mean, color = candidate_name),
    size = 1.2,  # Make the line thicker
    inherit.aes = FALSE
  ) +
  geom_ribbon(
    data = pred_summary,
    aes(x = end_date, ymin = pred_lower, ymax = pred_upper),
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  labs(
    x = "End Date",
    y = "Percentage",
    title = "Predicted Percentage Over Time",
    color = "Candidate",
    shape = "State"
  ) +
  theme_minimal() +
  theme(
    legend.key.size = unit(1, "cm"),
    legend.text = element_text(size = 8) 
  )

@@

## Overview of the Paper
This study employs a Bayesian modeling approach to forecast the 2024 U.S. presidential election, focusing on the candidates Kamala Harris and Donald Trump. By analyzing a comprehensive dataset of high-quality polls, the paper identifies the dynamics of voter support and how various factors—such as polling methodology, sample size, and geographic location—influence electoral outcomes. The findings indicate a greater likelihood of victory for Harris, projecting her with 222 electoral votes compared to Trump’s 109.

@@

## Objective and Overview
This survey methodology aims to forecast the U.S. presidential election outcome by gathering representative data from a diverse cross-section of American voters. The target population includes all eligible U.S. voters, spanning a wide array of demographic backgrounds. With a budget of $100,000, this methodology employs probability sampling techniques, effective recruitment, and thorough data validation protocols to ensure data accuracy and minimize bias. By applying probability sampling, each individual has a known chance of selection, allowing the results to be statistically valid and generalizable. This design accounts for demographic, geographic, and political influences, enhancing prediction reliability [@fivethirty; @pewresearch].

@@

# Additional data details

# Model details {#sec-model-details}

@@

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

>>>>

```{r}
#| include: false
#| warning: false
#| message: false
library(dplyr)
library(tidyr)
library(knitr)
library(rstanarm)
library(ggplot2)
library(tidyverse)
library(arrow)
library(janitor)
library(lubridate)
library(broom)
library(modelsummary)
library(splines)
library(caret)
```

@@

The analysis utilizes FiveThirtyEight's dataset of national presidential general election polls [@fivethirtyeight]. Following the approach outlined in [@tellingstories], we aim to predict the election outcome based on this polling data. The analyses were conducted in R [@citeR], utilizing several packages to streamline the process. The tidyverse packages [@citetidyverse] were essential for data simulation and preliminary testing. Initially, the raw data was downloaded using the arrow package [@citearrow], which facilitated efficient handling of large datasets.

Data cleaning employed dplyr [@citedplyr], tidyr [@citetidyr], and janitor [@citejanitor] to ensure consistent data structure, particularly for renaming columns and removing inconsistencies. Date variables were managed with the lubridate package [@citelubridate], making it easier to handle time-based data accurately. For modeling, rstanarm [@citerstanarm] was used for Bayesian regression modeling, while splines [@citesplines] handled non-linear relationships, and caret [@citecaret] supported model training and validation. The modelsummary [@citemodelsummary] package was used to summarize results, and broom [@citebroom] helped tidy up model outputs. Lastly, ggplot2 [@citeggplot] generated clear visualizations, ensuring effective communication of the findings.

@@

## Predictors Explanation
- **end date**: The date the poll concluded, marking the end of data collection for that specific poll.In the context of our analysis, we have set a criterion whereby the end dates must be greater than July 21, 2024. This threshold ensures that we are only considering the most recent and relevant polling data as we approach the 2024 election cycle.

@@

- **state**: The U.S. state where the poll was conducted, which allows for state-by-state analysis and comparison of support levels. @fig-state indicates the distribution of states.The histogram shows the distribution of polling data across different states. States like California and Georgia have high counts, indicating overrepresentation, while states such as Wyoming and South Dakota are underrepresented.

```{r}
#| label: fig-state
#| fig-cap: Distribution of State
#| echo: false
#| warning: false
#| message: false
#| fig-width: 10
#| fig-height: 6

ggplot(president_polls_cleaned_data, aes(x = state)) +
  geom_bar(fill = "pink") +
  labs(x = "State",
       y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Adjust text angle for better readability
```

- **sample size**: The number of respondents in the poll, indicating the scope and potential statistical reliability of the results. @fig-samplesize shows that smaller sample sizes (0 to 500) are significantly more common, indicating a high concentration of data points in this range and these specific sizes are frequently represented in the polling data, while larger sample sizes are relatively scarce.
```{r}
#| label: fig-samplesize
#| fig-cap: Distribution of Sample Size
#| echo: false
#| warning: false
#| message: false
#| fig-width: 6
#| fig-height: 3

# Load necessary libraries
library(ggplot2)
library(dplyr)

# Create a summary of counts for each sample_size
sample_size_summary <- president_polls_cleaned_data %>%
  group_by(sample_size) %>%
  summarise(count = n())

# Create a line plot of sample_size
ggplot(sample_size_summary, aes(x = sample_size, y = count)) +
  geom_line(color = "pink") +
  geom_point(color = "lightblue") +  # Optional: add points to highlight each count
  theme_minimal() +
  labs(#title = "Line Plot of Sample Size Counts",
       x = "Sample Size",
       y = "Count") 
#+ theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Adjust text angle for better readability

@@

@@

# Visualization
ggplot(president_polls_cleaned_data, aes(x = end_date, y = pct, color = candidate_name)) +
  geom_point(aes(shape = state), alpha = 0.5) +  # Use state as shape
  geom_line(
    data = pred_summary,
    aes(x = end_date, y = pred_mean, color = candidate_name),
    size = 1.2,  # Make the line thicker
    inherit.aes = FALSE
  ) +
  geom_ribbon(
    data = pred_summary,
    aes(x = end_date, ymin = pred_lower, ymax = pred_upper),
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  labs(
    x = "End Date",
    y = "Percentage",
    #title = "Predicted Percentage Over Time",
    color = "Candidate",
    shape = "State"
  ) +
  theme_minimal() +
  theme(
    legend.key.size = unit(1, "cm"),
    legend.text = element_text(size = 8) 
  )

@@

## Overview of the Paper
This study employs a Bayesian modeling approach to forecast the 2024 U.S. presidential election, focusing on candidates Kamala Harris and Donald Trump. By analyzing a robust dataset of high-quality polls, the research investigates how factors such as polling methodology, sample size, and geographic location influence voter support. The findings indicate that Harris is projected to secure 222 electoral votes, while Trump is expected to receive 109.

These results reflect broader social dynamics that shape electoral behavior. Variations in support across states suggest that regional identities and socio-political contexts significantly influence public opinion. Research indicates that diverse populations often favor candidates advocating for progressive policies, which align with issues such as racial equity, healthcare access, and economic opportunity (@pew2020). Conversely, more conservative regions tend to support traditional Republican platforms, reflecting local values and priorities (@brookings2020).

@@

## Objective and Overview
This survey methodology aims to forecast the U.S. presidential election outcome by gathering representative data from a diverse cross-section of American voters. The target population includes all eligible U.S. voters, spanning a wide array of demographic backgrounds. With a budget of $100,000, this methodology employs probability sampling techniques, effective recruitment, and thorough data validation protocols to ensure data accuracy and minimize bias. By applying probability sampling, each individual has a known chance of selection, allowing the results to be statistically valid and generalizable. This design accounts for demographic, geographic, and political influences, enhancing prediction reliability [@fivethirtyeight; @pewresearch].

@@

\newpage
# Model details {#sec-model-details}

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheck
#| fig-width: 6
#| fig-height: 4
#| fig-cap: "Posterior Predictive Check of Bayesian Models"

# Load the model using the here function to specify the path
model_bayes <- readRDS("../models/election_glm_model.rds")
# Perform posterior predictive check
pp_check(model_bayes)
```

@fig-ppcheck presents a posterior predictive check (PPC) for Bayesian model, illustrating the relationship between the observed data and the data replicated by the model. On the x-axis, the values of the response variable are depicted, while the y-axis indicates the density or probability of those values. The dark blue curve represents the observed data, denoted as $y$, and is contrasted with multiple light blue curves that signify the model's replicated data, referred to as $y_rep$. 
The close overlap between the dark blue curve and the cluster of light blue curves suggests that the model fits the observed data well, indicating its ability to replicate the distribution of the data accurately. Additionally, the variability among the light blue curves highlights the uncertainty in the model's predictions; if these curves were excessively dispersed or showed minimal overlap with the observed data, it would signal potential issues with the model's fit.

## Diagnostics
### R-hat Plots

$\hat{R}$ is a diagnostic measure that compares the variance within each chain to the variance between chains. @fig-rhat presents the R-hat values for Bayesian model, which serves as a diagnostic indicator for evaluating the convergence of Markov Chain Monte Carlo (MCMC) chains. The x-axis indicates the R-hat statistic, reflecting the convergence status of each parameter in the model, while the y-axis lacks a specific scale and displays various R-hat values along a vertical line. Each point represents the R-hat value for a different parameter, and their clustering around 1.00 suggests effective convergence.
Additionally, @fig-rhat includes interpretation thresholds for R-hat values: values at or below 1.05 signify satisfactory convergence, while values up to 1.1 may indicate some potential convergence issues but are generally considered acceptable. In contrast, values exceeding 1.1 suggest poor convergence, indicating that further iterations may be necessary to obtain stable estimates. Since most of the R-hat values in @fig-rhat are well below 1.05, we can infer that the Bayesian model that we construnct is likely well-fitted and does not raise significant concerns regarding the convergence of the MCMC process.

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-rhat
#| fig-cap: "R-hat Plots of Bayesian Model"
#| fig-width: 6
#| fig-height: 4

plot(model_bayes, "rhat")

### Trace Plots

Trace plot of  Bayesian model illustrates the sampled values of various parameters across multiple iterations of the Markov Chain Monte Carlo (MCMC) process, and discusses the convergence of the MCMC chains. Organized in a grid format, each row represents a different parameter, while each column corresponds to a specific MCMC chain. The x-axis indicates the iteration number of the MCMC sampling, and the y-axis shows the values of the parameters being sampled, with different colors or line styles potentially representing various MCMC chains. Ideally, the sampled values should fluctuate around a stable mean, indicating effective convergence. Conversely, if any parameter shows a lack of mixing or clear trends, such as drifting over time, it may indicate convergence issues, necessitating additional iterations or adjustments to improve the model.

@fig-trace presents a trace plot for the Bayesian model which consider different states, variable end date, variable sample size and variable candidate name. We can see that there is no clear separation between the chains and sampled values fluctuate around the stable mean, suggesting the MCMC sampling has successfully converged for most parameters.

```{r}
#| echo: false
#| eval: true
#| label: fig-trace
#| fig-cap: "Trace Plot of Bayesian Model"
#| warning: false
#| fig-width: 12
#| fig-height: 12

plot(model_bayes, "trace")
```
